Exploring environmental impacts of colonial contacts on Indigenous communities in northeastern Taiwan

### 2024-07-21 analysis/paper/paper.qmd
revised paragraphs

# Introduction
<!-- introduction about European expansion -->
European expansion since the 15th century marked the beginning of an era characterized by global trade, exploration, and colonization, which often led to profound impacts on indigenous societies in many parts of the world. The forms of European colonization and power relations varied across regions, depending on specific spatial and temporal contexts [@Jorden2009]. Understanding the dynamics of these colonial processes has become a significant issue in archaeology, as it provides critical insights into the complex outcomes of European colonization [@Gosden2004]. In Asia, European colonization was primarily driven by the procurement of Asian luxury goods, raw materials, or natural resources for commercial profit. The new exchange systems and governance policies imposed by European colonizers often had profound effects on local economic systems, social structures, and settlement patterns [@Christie1996; @Junker2004; @Lape2002]. However, Indigenous communities did not just passively accept European colonization. Recent discussions on colonial interactions also highlight the diverse interactions employed by Indigenous communities in response to European presence, such as resistance, negotiation, persistence, resilience, and entanglement [@Torrence2003; @Silliman2005; @Mitchell2010; @Jorden2009; @Alexander1998].

@@

# Colonial encounter in Northeastern Taiwan
<!-- the structure of power relations, colonial control in particular spatial and temporal contexts-->
In 1626, the Spanish colonized northern Taiwan and built fortress in Tamsui and Keelung to the threat of the VOC (Vereenigde Oost-Indische Compagnie; Dutch East India Company) in Asian market, which established a colony in southern Taiwan two years before the Spanish in the north. In 1642, the Dutch defeated the Spanish and occupied their fortresses, officially expanding their colonial power to northern Taiwan until the end of their colonization in Taiwan in 1662 [@Andrade2007]. The form of European colonization in northern Taiwan relied to a certain degree on local Indigenous people for the development of the colony and resources collections among different Indigenous groups. For example, the Basay people, an indigenous group active in Keelung, were recorded to have assisted in the colonial activities of the VOC administrators, such as coordinating between indigenous villages [@Kang2016]. The colonial domination was mostly established through the conquest over indigenous villages, but the colonial control varied depending on the location of indigenous settlements. For most indigenous villages distant from the European colonial base, the colonial control was relatively less, indirect, and even unsuccessful.
<!-- domination and the processes by which it was established and resisted, how and why colonial powers were able to exert control (however limited) over Native populations -->

>>>>

# Introduction
<!-- introduction about European expansion -->
European expansion since the 15th century marked the beginning of an era characterized by global trade, exploration, and colonization, which often led to profound impacts on indigenous societies in many parts of the world. The forms of European colonization and power relations varied across regions, depending on specific spatial and temporal contexts [@Jorden2009]. Understanding the dynamics of these colonial processes has become a significant issue in archaeology, as it provides critical insights into the complex outcomes of European colonization [@Gosden2004]. In Asia, European colonization was primarily driven by the procurement of Asian luxury goods, raw materials, or natural resources based on commercial interest [@Prakash1997]. The new exchange systems and governance policies imposed by European colonizers often had profound effects on local economic systems, social structures, and settlement patterns [@Christie1996; @Junker2004; @Lape2002]. However, Indigenous communities did not just passively accept European colonization. Recent discussions on colonial interactions also highlight the diverse interactions employed by Indigenous communities in response to European presence, such as resistance, negotiation, persistence, resilience, and entanglement [@Torrence2003; @Silliman2005; @Mitchell2010; @Jorden2009; @Alexander1998].

@@

# Colonial encounter in Northeastern Taiwan
<!-- the structure of power relations, colonial control in particular spatial and temporal contexts-->
In 1626, the Spanish arrived Keelung in northern Taiwan and built Fort San Salvador at Heping Dao and later Fort Santo Domingo in Tamsui to counter the threat of the VOC (Vereenigde Oost-Indische Compagnie; Dutch East India Company) in Asian market, which had established a colony in southern Taiwan two years earlier [@Borao2009, p.41; @Andrade2007]. Despite an attempt to build a colonial city, San Salvador remained in its early stages of development and functioned as a presidio (military outpost), hosting mainly Spanish officers, soldiers, and missionaries [@Borao2009, p.103-134]. The area outside of Fort San Salvador was inhabited by local indigenous groups, the Taparris and Quimaurris, who had resided at the location of the fort before being expelled by the Spanish [@Borao2001, SIT:165, p.81]. The Spanish colonization may not have involved effective colonialism or asymmetrical power relations. Instead, the Spanish had to maintain good relations with the local indigenous people, as they relied on them for food supply to a certain degree [@Borao2001, SIT:170-171]. Also, their territorial control could have been limited to a small area around the fortress, with minimal indigenous involvement only in exchange of goods.

In 1642, the Dutch defeated the Spanish and took over their fortresses, officially expanding their colonial power to northern Taiwan until the end of the colonization in 1662. Compared with the Spanish, the Dutch were more strategic in consolidating their colonial power by forming alliances with the natives, imposing a bureaucratic and tribute system, and conquering indigenous villages that were unwilling to comply [@Kang2016; @Andrade2007]. Both Spanish and Dutch worked with local indigenous people for the development of the colony, especially for the Dutch. For example, the Basay people active in Keelung, was recorded to have assisted in coordinating or collecting tribute among different Indigenous groups for the VOC governors and they also worked with the Spanish governors in early time [@Kang2016]. The colonial domination was mostly established through the conquest over indigenous villages and offering protection to the indigenous people, but the colonial control varied depending on the location of indigenous settlements. For most indigenous villages distant from the European colonial base, the colonial control was relatively less, indirect, and even unsuccessful.

<!-- domination and the processes by which it was established and resisted, how and why colonial powers were able to exert control (however limited) over Native populations -->

A church was built later that reveals their religious activities [@Berrocal2018].
